Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-02 notebooks/expected_and_actual_counts_ratio.Rmd
Plot histogram of p-values

library(data.table)
library(ggplot2)
library(stringr)
```

@@

# Create a list of the clades to include in comparisons.
clades <- c("22A","22B")

@@

```{r filter_all_data}
# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)

# Add column with the mutation type (clade founder.mutation).
#filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
```

@@

@@

```{r stacked_bar, echo=FALSE}
# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
#ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  #geom_bar(position="stack", stat="identity")+
  #theme_bw()+
  #labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```

@@

```{r filter_contingency_data}
# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_mutation, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$rounded_expected, observed_expected=clade_SARS_mutation$observed_expected)
```

@@

```{r combined_clade_comparisons}
# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_SARS_mutation$nt_mutation), each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind", replicate(length(unique(filtered_SARS_mutation$nt_mutation)), clade_comparisons, simplify=FALSE))

@@

```{r filter_for_comparisons}
# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
filtered_comparisons <- comparisons[(comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade),]
```

@@

# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
clade1 <- merge(filtered_comparisons[,c("id","clade1")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade1", by.y="nt_mutation_clade")
clade2 <- merge(filtered_comparisons[,c("id","clade2")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade2", by.y="nt_mutation_clade")

# Add observed, expected and observed/expected data to the comparison matrix.
{filtered_comparisons$clade1_observed <- clade1[order(clade1$id),"observed"]
filtered_comparisons$clade1_expected <- clade1[order(clade1$id),"expected"]
filtered_comparisons$clade2_observed <- clade2[order(clade2$id),"observed"]
filtered_comparisons$clade2_expected <- clade2[order(clade2$id),"expected"]
filtered_comparisons$clade1_obs_exp <- clade1[order(clade1$id),"observed_expected"]
filtered_comparisons$clade2_obs_exp <- clade2[order(clade2$id),"observed_expected"]}
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))

@@

```{r combine_p_values}
# Combine p-value list with the clade comparison list.
comparison_p_values <- cbind(filtered_comparisons, p_values)
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
# Double check calculation with Jesse's paper
comparison_p_values$fold_change <- log2(ifelse(comparison_p_values$clade1_obs_exp >= comparison_p_values$clade2_obs_exp, comparison_p_values$clade1_obs_exp/comparison_p_values$clade2_obs_exp, -comparison_p_values$clade2_obs_exp/comparison_p_values$clade1_obs_exp))
```

@@

```{r volcano_plot}
# Only plot data for the two clades being compared.
#plot_data <- comparison_p_values[which(comparison_p_values$clades=="20A.22A"),]
ggplot(comparison_p_values, aes(x=other_fold, y=-log10(p_values_corrected)))+
  geom_point()
```

@@

```{r verify_analyses}
# sanity check analyses (distribution of synonymous p-values)

```

## Shorted Script That Can Handle 2 Clades
```{r two_clades}
# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds
{start_two <- Sys.time()
SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
country <- "all"
clades <- c("20A","20B")
min_expected <- 0
pseudo <- 0.5
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
clade1 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[1]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
clade2 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[2]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
m <- merge(clade1, clade2, by="nt_mutation")
filtered_comparisons <- data.frame(clade1=paste(m$nt_mutation, m$clade.x, sep="_"), clade2=paste(m$nt_mutation, m$clade.y, sep="_"), clade1_observed=m$actual_count.x, clade1_expected=m$expected_count.x, clade2_observed=m$actual_count.y, clade2_expected=m$expected_count.y, clade1_obs_exp=m$observed_expected.x, clade2_obs_exp=m$observed_expected.y)
end_two <- Sys.time()}
end_two - start_two
two <- filtered_comparisons
```

>>>>

library(data.table)
library(ggplot2)
library(stringr)
library(gtools)
```

@@

# Create a list of the clades to include in comparisons.
clades <- c("20A","22B")

@@

```{r filter_all_data}
# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
```

@@

The Fisher's Exact test can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).

@@

```{r stacked_bar, echo=FALSE}
# Add column with the mutation type (clade founder.mutation).
#filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")

# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
#ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  #geom_bar(position="stack", stat="identity")+
  #theme_bw()+
  #labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```

@@

```{r filter_contingency_data}
# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
filtered_clades$nt_mutation_clade <- paste(filtered_clades$nt_mutation, filtered_clades$clade, sep="_")
```

@@

```{r combined_clade_comparisons}
# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_clades$nt_mutation), each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind", replicate(length(unique(filtered_clades$nt_mutation)), clade_comparisons, simplify=FALSE))

@@

```{r filter_for_comparisons}
# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
filtered_comparisons <- comparisons[(comparisons$clade1 %in% filtered_clades$nt_mutation_clade & comparisons$clade2 %in% filtered_clades$nt_mutation_clade),]
```

@@

# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
clade1 <- merge(filtered_comparisons[,c("id","clade1")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade1", by.y="nt_mutation_clade")
clade2 <- merge(filtered_comparisons[,c("id","clade2")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade2", by.y="nt_mutation_clade")

# Change column names so that clade 1 and clade 2 data frames can be combined
colnames(clade1)[3:ncol(clade1)] <- paste("clade1", colnames(clade1)[3:ncol(clade1)], sep="_")
colnames(clade2)[3:ncol(clade2)] <- paste("clade2", colnames(clade2)[3:ncol(clade2)], sep="_")

# Order clade 1 and clade 2 data frames by id and combine
fisher_data <- cbind(clade1[order(clade1$id),], clade2[order(clade2$id),])
fisher_data[,which(colnames(fisher_data) == "id")] <- NULL
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
'p_values <- data.frame(p_value=apply(fisher_data, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'

@@

```{r combine_p_values}
# Combine p-value list with the clade comparison list.
comparison_p_values <- cbind(fisher_data, p_values)
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
# Double check calculation with Jesse's paper
comparison_p_values$fold_change <- log2(ifelse(comparison_p_values$clade1_observed_expected >= comparison_p_values$clade2_observed_expected, comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected, -comparison_p_values$clade2_observed_expected/comparison_p_values$clade1_observed_expected))
```

@@

```{r volcano_plot}
# Only plot data for the two clades being compared.
#plot_data <- comparison_p_values[which(comparison_p_values$clades=="20A.22A"),]
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_values_corrected)))+
  geom_point()
```

@@

```{r verify_analyses}
# sanity check analyses (distribution of synonymous p-values)
plot_data <- comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),]
nb <- sqrt(nrow(plot_data))
bw <- (max(comparison_p_values$p_values_corrected)-min(comparison_p_values$p_values_corrected))/nb
ggplot(plot_data, aes(x=p_values_corrected))+
  geom_histogram(bins = nb, binwidth = bw)

ggplot(comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),], aes(x=p_values_corrected, y=p_values_corrected))+
  geom_point()
```

## Shorted Script That Can Handle 2 Clades
```{r two_clades}
# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds
'{start_two <- Sys.time()
SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
country <- "all"
clades <- c("20A","20B")
min_expected <- 0
pseudo <- 0.5
filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
clade1 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[1]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
clade2 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[2]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
m <- merge(clade1, clade2, by="nt_mutation")
filtered_comparisons <- data.frame(clade1=paste(m$nt_mutation, m$clade.x, sep="_"), clade2=paste(m$nt_mutation, m$clade.y, sep="_"), clade1_observed=m$actual_count.x, clade1_expected=m$expected_count.x, clade2_observed=m$actual_count.y, clade2_expected=m$expected_count.y, clade1_obs_exp=m$observed_expected.x, clade2_obs_exp=m$observed_expected.y)
end_two <- Sys.time()}
end_two - start_two
two <- filtered_comparisons'
```
